Best Practices for Alchemical Free Energy Calculations: v\versionnumber

### 2020-01-20 paper/manuscript.tex
trying to fiddle about with figure placement and checklist

    \includegraphics[width=0.95\linewidth]{paper/figures/fig5_thermo_cyc_abs/Figure.pdf}

@@

Another way of stating this is that intermediate states should sample molecular configurations that are as similar as possible to their neighboring states. The more similar the configurations are between intermediate states, the lower the statistical uncertainty is in the estimate of free energy between intervals. This can be proven directly from the BAR and MBAR formulas~\cite{bennett1976efficient,klimovich2015guidelines}, though the exact same principles apply for TI. For a 'good' path to work sufficient overlap between pontential energies is required. Fig.~\ref{fig:fig_what_is_lambda} (\textbf{A}) and (\textb{B}) try to illustrate this. Fig.~\ref{fig:fig_what_is_lambda} (\textbf{A}) shows in a pictorial way a soft-core potential can be applied across different $\lamda$s. Fig.~\ref{fig:fig_what_is_lambda} (\textbf{B}) illustrates the potential energy distributions at the different $\lambda$ intermediates, with sufficient overlap between neighboring $\lambda$ states to ensure that reweighting estimators such as MBAR can be used for analysis (see Sec.\ref{subsec:estimators}). The actual transformation is best handled with soft-core potentials of the form shown in Fig.\ref{fig:fig_what_is_lambda} (\textbf{C}) and (\textbf{B}), with more details given below. 

@@

One way of assessing reliability of the calculations is checking the phase space overlap between neighboring $\lambda$-windows~\cite{wu2005phasespaceb, wu2005phasespacec}. For this purpose, a so-called overlap matrix $\mathcal{O}$ can be used.  $\mathcal{O} \in \mathbb{R}^{K\cross K}$, with $K$ being the number of simulated states, i.e. $\lambda$s. Sufficient overlap is important for reweighting estimators such as BAR or MBAR, but cannot help assess reliability of estimates, when using TI. 

@@

Interestingly, as the variance is inversely correlated with the number of states\cite{klimovich2015guidelines}, it can in principle be reduced below any arbitrary threshold with enough simulation time and a large enough number of $\lambda$ windows. However, decreasing the variance to a value close to 0 is not feasible, as this approach would strongly increase the calculation time. While variance can be decreased both by increasing simulation length and by increasing the number of $\lambda$ values, if overlap is poor, the latter approach often has a larger immediate impact.  Specifically, often the goal is to adjust the $\lambda$ spacing to perform additional sampling in the areas of poor overlap, rather than using arbitrarily spaced windows. Different approaches are described in section~\ref{sec:simulation_protocol_choice}) and more details can be found in the literature~\cite{dakka2018concurrent, hahn2019alchemical}.

@@

\section*{Author Contributions}
%%%%%%%%%%%%%%%%
% This section mustt describe the actual contributions of
% author. Since this is an electronic-only journal, there is
% no length limit when you describe the authors' contributions,
% so we recommend describing what they actually did rather than
% simply categorizing them in a small number of
% predefined roles as might be done in other journals.
%
% See the policies ``Policies on Authorship'' section of https://livecoms.github.io
% for more information on deciding on authorship and author order.
%%%%%%%%%%%%%%%%
\todo[inline, color={green!20}]{ASJSM: @everyone, can you check your contribution is acknowledge and if not add the appropriate information. }
(Explain the contributions of the different authors here)
\textbf{ASJSM}: Coordinated the document, contributed to most sections, and co-designed figures ~\ref{} and created figures \ref{fig:fig_what_is_lambda},~\ref{fig:fig1_what_is_alchemy},~\ref{fig:overlap} and replotted~\ref{fig:automatic-equilibration-detection} and improved ~\ref{fig:fig_types_of_networks}.\\
\textbf{JS}: created figures \ref{fig:fig_what_is_alchemy},~\ref{fig:fig_binding_thermodynamic_cycle},~\ref{fig:fig_topology},~\ref{fig:fig_mcss},~\ref{fig:fig_types_of_networks},~\ref{fig:fig_absolute_thermodynamic_cycle},~\ref{fig:scatterplot_analysis}, the data representation section, the checklist and contributed to general formatting discussions and editing.\\
\textbf{MK}: Contributed to the Data Analysis section, provided the data for figure~\ref{fig:overlap} and helped edit the paper.\\
DLM: Contributed to the outline, drafted some of the sections, gave ideas on figures, and helped edit the paper.\\
\textbf{GT}: Contributed to introduction and drug discovery sections, and helped edit the paper.\\
\textbf{AR}: Created figure~\ref{fig:freeenergytrajectories}, contributed to section \ref{sec:simulation_protocol_choice}, and helped edit the paper.\\
\textbf{MRS}: Helped create figure~\ref{fig:fig_what_is_lambda}, wrote sections describing choices for alchemical pathways and analysis for free energy calculations.\\
\textbf{LNN}: Helped write the simulation length, stopping conditions, and information saving section.\\
\textbf{BA}: Helped write the uncertainty estimation, stopping conditions, and output analysis sections and created figure ~\ref{fig:convergence_forward_reverse}.
\textbf{JM}: Contributed to sections~\ref{subsec:reproducible},~\ref{sec:prerequisites},~\ref{} 4.2, 5.3, 6, 8.3.1, 8.3.2\\
\textbf{JDC}: Wrote section~\ref{sec:decorrelating-samples} and~\ref{sec:automatic-equilibration-detection} discussed structure and design of the whole document, suggested Figs.~\ref{fig:fig_what_is_alchemy} and~\ref{fig:fig_sampling_scheme} \todo[inline, color={red!20}]{@JDC check info}\\
\textbf{HX}: Contributed section~\ref{subsec:accuracy}, to section~\ref{sec:relative-fe-protocol}, and to section~\ref{sec:are-they-good}.

% We suggest you preserve this comment:
For a more detailed description of author contributions,
see the GitHub issue tracking and changelog at \githubrepository.

>>>>

    \includegraphics[width=0.95\linewidth]{figures/fig5_thermo_cyc_abs/Figure.pdf}

@@

Another way of stating this is that intermediate states should sample molecular configurations that are as similar as possible to their neighboring states. The more similar the configurations are between intermediate states, the lower the statistical uncertainty is in the estimate of free energy between intervals. This can be proven directly from the BAR and MBAR formulas~\cite{bennett1976efficient,klimovich2015guidelines}, though the exact same principles apply for TI. For a 'good' path to work sufficient overlap between pontential energies is required. Fig.~\ref{fig:fig_what_is_lambda} (\textbf{A}) and (\textbf{B}) try to illustrate this. Fig.~\ref{fig:fig_what_is_lambda} (\textbf{A}) shows in a pictorial way a soft-core potential can be applied across different $\lambda$s. Fig.~\ref{fig:fig_what_is_lambda} (\textbf{B}) illustrates the potential energy distributions at the different $\lambda$ intermediates, with sufficient overlap between neighboring $\lambda$ states to ensure that reweighting estimators such as MBAR can be used for analysis (see Sec.\ref{subsec:estimators}). The actual transformation is best handled with soft-core potentials of the form shown in Fig.\ref{fig:fig_what_is_lambda} (\textbf{C}) and (\textbf{B}), with more details given below. 

@@

One way of assessing reliability of the calculations is checking the phase space overlap between neighboring $\lambda$-windows~\cite{wu2005phasespaceb, wu2005phasespacec}. For this purpose, a so-called overlap matrix $\mathcal{O}$ can be used.  $\mathcal{O} \in \mathbb{R}^{K\times K}$, with $K$ being the number of simulated states, i.e. $\lambda$s. Sufficient overlap is important for reweighting estimators such as BAR or MBAR, but cannot help assess reliability of estimates, when using TI. 

@@

Interestingly, as the variance is inversely correlated with the number of states~\cite{klimovich2015guidelines}, it can in principle be reduced below any arbitrary threshold with enough simulation time and a large enough number of $\lambda$ windows. However, decreasing the variance to a value close to 0 is not feasible, as this approach would strongly increase the calculation time. While variance can be decreased both by increasing simulation length and by increasing the number of $\lambda$ values, if overlap is poor, the latter approach often has a larger immediate impact.  Specifically, often the goal is to adjust the $\lambda$ spacing to perform additional sampling in the areas of poor overlap, rather than using arbitrarily spaced windows. Different approaches are described in section~\ref{sec:simulation_protocol_choice}) and more details can be found in the literature~\cite{dakka2018concurrent, hahn2019alchemical}.

@@

\section{Checklist}
\label{sec:checklist}
\begin{Checklists*}[p!]
\begin{checklist}{ Know what you want to simulate}
    \textbf{Initial questions you should ask before you set up an alchemical free energy calculation using molecular dynamics simulations}
\begin{itemize}
    \item Do I understand the biology, chemistry and physics of my system?
    \item Have I properly prepared my protein and ligand systems?
    \item Does my system contain any structures that require custom parameters?
    \item What simulation protocol will provide the most evidence to answer my hypothesis?
    \item Are the projected computational expense and runtime realistic for my goals?
    \item Will my protocol be reproducible? 
    \item Will my statistics be reliable? If not, would more replicates solve the problem? Three replicates is the minimal number of replicates that should be run
    \item Can I open-source my data?
\end{itemize}
\end{checklist}

\begin{checklist}{Preparing your simulations}
\textbf{Steps to getting started setting up your alchemical free energy calculation}
\begin{itemize}
    \item Make sure you know why you have picked your (combination of) force field(s)
    \item Minimize your system
    \item Equilibrate your system with your choice of thermodynamic ensemble
    \item Check the stability of your system and whether it behaves the way you want it to
\end{itemize}
\end{checklist}

\begin{checklist}{Running absolute simultions}
        \textbf{Steps to running your absolute alchemical free energy calculations}
\begin{itemize}
 \item Your ligands should have the same, biologically correct binding pose
        \item Make sure your \textlambda-scheduling is set correctly
        \item Check if your ligands are discharging and vanishing correctly
        \item Set up your restraints correctly
        \item Make sure you subsample in your MBAR protocol
        \item Apply the right correction terms
\end{itemize}
\end{checklist}

\begin{checklist}{Running relative simulations}
        \textbf{Steps to running your relative alchemical free energy calculations}
\begin{itemize}
   \item Your ligands should have the same, biologically correct binding pose
        \item Make sure your λ-scheduling is set correctly
        \item Make sure your molecular transformations are realistic (1-5 heavy atoms for reliable computations)
        \item Generate a perturbation network by your method of choice; check whether you have enough cycle closures (preferably, every ligand should be in a cycle)
        \item Check whether dummy atoms were assigned correctly
        \item Make sure you subsample in your MBAR protocol
\end{itemize}
\end{checklist}

\begin{checklist}{How do I know which simulations are not reliable}
    \textbf{Situations suggesting your relative alchemical free energy calculations have not run properly (assuming absence of experimental affinities)}
        \begin{itemize}
                \item Standard error (\textsigma) should be \textless1 kcal·mol-1 
    \item Simulated systems have not converged 
    \newline\newline\textit{Relative:}
    \item If you observe hysteresis in perturbations and incorrect cycle closures
    \item Energy differences \textgreater15 kcal·mol-1  are unreliable
    \newline\newline\textit{Absolute:}
    \item Energies below -15 kcal·mol-1  are unreliable
    \item The ligand has not sampled most of the intended region after the vanishing step
    \item The ligand is drifting out of the intended region after the vanishing step
        \end{itemize}
\end{checklist}

\begin{checklist}{Why are they not reliable}
    \textbf{Suggestions for finding out why your alchemical free energy calculations may not be reliable}
\begin{itemize}
    \item Check again whether dummy atoms were assigned correctly
    \item Inspect the trajectories across the λ-schedule (particularly the endpoints)
    \item Inspect the overlap matrices
\end{itemize}
\end{checklist}

\begin{checklist}{Data Analysis}
    \textbf{Steps to analyzing your output data correctly}
\begin{itemize}
    \item Make sure you have run enough replicates to ensure statistical reliability (\textgreater3)
    \item Compute both correlation and ranking coefficients and ranking statistics (e.g. r, \textrho, MUE and \texttau)
    \item You should be able to explain large energy differences based on your system (trajectory)
\end{itemize}
\end{checklist}
\end{Checklists*}
\clearpage

\section*{Author Contributions}
%%%%%%%%%%%%%%%%
% This section mustt describe the actual contributions of
% author. Since this is an electronic-only journal, there is
% no length limit when you describe the authors' contributions,
% so we recommend describing what they actually did rather than
% simply categorizing them in a small number of
% predefined roles as might be done in other journals.
%
% See the policies ``Policies on Authorship'' section of https://livecoms.github.io
% for more information on deciding on authorship and author order.
%%%%%%%%%%%%%%%%
\todo[inline, color={green!20}]{ASJSM: @everyone, can you check your contribution is acknowledge and if not add the appropriate information. }
(Explain the contributions of the different authors here)
\textbf{ASJSM}: Coordinated the document, contributed to most sections, and co-designed figures ~\ref{} and created figures \ref{fig:fig_what_is_lambda},~\ref{fig:fig1_what_is_alchemy},~\ref{fig:overlap} and replotted~\ref{fig:automatic-equilibration-detection} and improved ~\ref{fig:fig_types_of_networks}.\\
\textbf{JS}: created figures \ref{fig:fig_what_is_alchemy},~\ref{fig:fig_binding_thermodynamic_cycle},~\ref{fig:fig_topology},~\ref{fig:fig_mcss},~\ref{fig:fig_types_of_networks},~\ref{fig:fig_absolute_thermodynamic_cycle},~\ref{fig:scatterplot_analysis}, the data representation section, the checklist and contributed to general formatting discussions and editing.\\
\textbf{MK}: Contributed to the Data Analysis section, provided the data for figure~\ref{fig:overlap} and helped edit the paper.\\
\textbf{DLM}: Contributed to the outline, drafted some of the sections, gave ideas on figures, and helped edit the paper.\\
\textbf{GT}: Contributed to introduction and drug discovery sections, and helped edit the paper.\\
\textbf{AR}: Created figure~\ref{fig:freeenergytrajectories}, contributed to section \ref{sec:simulation_protocol_choice}, and helped edit the paper.\\
\textbf{MRS}: Helped create figure~\ref{fig:fig_what_is_lambda}, wrote sections describing choices for alchemical pathways and analysis for free energy calculations.\\
\textbf{LNN}: Helped write the simulation length, stopping conditions, and information saving section.\\
\textbf{BA}: Helped write the uncertainty estimation, stopping conditions, and output analysis sections and created figure ~\ref{fig:convergence_forward_reverse}.\\
\textbf{JM}: Contributed to sections~\ref{subsec:reproducible},~\ref{sec:prerequisites},~\ref{} 4.2, 5.3, 6, 8.3.1, 8.3.2\\
\textbf{JDC}: Wrote section~\ref{sec:decorrelating-samples} and~\ref{sec:automatic-equilibration-detection} discussed structure and design of the whole document, suggested Figs.~\ref{fig:fig_what_is_alchemy} and~\ref{fig:fig_sampling_scheme} \\
\textbf{HX}: Contributed section~\ref{subsec:accuracy}, to section~\ref{sec:relative-fe-protocol}, and to section~\ref{sec:are-they-good}.\\
% We suggest you preserve this comment:
For a more detailed description of author contributions,
see the GitHub issue tracking and changelog at \githubrepository.
